---
title: "NCAA-Analysis"
output: html_document
date: "2022-10-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



We download and load necessary packages below
```{r}
#install.packages("lmerTest")
#install.packages("sjPlot")
#install.packages("estimatr")
#install.packages('car')
#install.packages("corrplot")
#install.packages("MASS")

library("lmerTest")
library("tidyr")
library("psycho")
library("sjPlot")
library("estimatr")
library("car")
library("corrplot")
library("MASS")
library("dplyr")

```


We import the filtered NCAA boxscore below
```{r}

box <- read.csv("/Users/chejeong/Desktop/FA2022/DA401/NBA-NCAA-analytics/Data/NCAA/ncaa_boxscore_filtered2.csv")

```

We change the dependent variable as a factor
```{r}

box$win <- as.factor(box$win)

```

We select relevant variables
```{r}

box <- box %>% dplyr::select(win,season,three_points_att,three_points_made,
                    two_points_att,two_points_made,free_throws_att,
                    free_throws_made,offensive_rebounds, 
                    defensive_rebounds, assists, 
                    turnovers, steals, blocks, personal_fouls, 
                    fast_break_pts, second_chance_pts, points_off_turnovers)

```


We filter out null values
```{r}

print(dim(box))
print(colSums(is.na(box)))

box <- box %>% drop_na()

#export
#write.csv(box,"/Users/chejeong/Desktop/FA2022/DA401/NBA-NCAA-analytics/Data/NCAA/ncaa_boxscore_filtered3.csv", row.names = FALSE)

```


### Mixed Effects Logistic Regression Assumption Test

We test the mixed effects logistic regression assumptions below:

Linearity: relationship between the natural log of these probabilities (when expressed as odds) and your predictor variable is linear.

No outliers: independent variables should not contain outliers

Multicollinearity: independent variables should not be substantially correlated to each other.

*Linearity*

We test the lineartiy of the data.


*Outliers*

We explore if there are any outliers in the independent variables of interest below.
                  
"three_points_made"    "three_points_att"    
[17] "three_points_pct"     "two_points_made"      "two_points_att"       "two_points_pct"      
[21] "free_throws_made"     "free_throws_att"      "free_throws_pct"      "offensive_rebounds"  
[25] "defensive_rebounds"   "rebounds"             "assists"              "turnovers"           
[29] "steals"               "blocks"               "personal_fouls"       "foulouts"            
[33] "points"               "fast_break_pts"       "second_chance_pts"    "points_off_turnovers"

```{r}

boxplot(box$three_points_made,ylab = "3PM")
boxplot(box$three_points_att,ylab = "3PA")
boxplot(box$two_points_made,ylab = "2PM")
boxplot(box$two_points_att,ylab = "2PA")
boxplot(box$free_throws_made,ylab = "FTM")
boxplot(box$free_throws_att,ylab = "FTA")
boxplot(box$offensive_rebounds,ylab = "OFR")
boxplot(box$defensive_rebounds,ylab = "DFR")
boxplot(box$assists,ylab = "AST")
boxplot(box$turnovers,ylab = "TO")
boxplot(box$steals,ylab = "STL")
boxplot(box$blocks,ylab = "BLK")
boxplot(box$fast_break_pts,ylab = "Fastbreak PTS")
boxplot(box$second_chance_pts,ylab = "Second Chance PTS")
boxplot(box$points_off_turnovers,ylab = "PTS off TO")

```

*Multicollinearity*

We test multicollinearity with the code below

```{r}

#fit linear model below
lm <- glm(win ~ three_points_att + three_points_made + two_points_att + two_points_made + free_throws_att + free_throws_made + offensive_rebounds + defensive_rebounds + assists + turnovers + steals + blocks + personal_fouls + fast_break_pts + second_chance_pts + points_off_turnovers,data=box, family = "binomial")

```

We test for multicollinearity using vif below.

```{r}

vif(lm)

```


```{r}

data_x <- box2 %>% dplyr::select(three_points_att, three_points_made, two_points_att,
                    two_points_made,free_throws_att,free_throws_made,
                    offensive_rebounds, defensive_rebounds, assists, 
                    turnovers, steals, blocks, personal_fouls, 
                    fast_break_pts, second_chance_pts, points_off_turnovers)


corrplot(cor(data_x), method = 'shade',is.corr=F, number.cex=0.3, addCoef.col="black")

```


### Model Fitting

We fit the Mixed Effects Logistic Regression below

```{r}

m <- glmer(win ~ three_points_att + three_points_made + two_points_att + free_throws_att + free_throws_made + offensive_rebounds + defensive_rebounds + assists + turnovers + steals + blocks + personal_fouls + fast_break_pts + second_chance_pts + points_off_turnovers + (1 | season),data=box, family = binomial,control = glmerControl(optimizer = "bobyqa"))

```

```{r}

summary(m)

```

```{r}
tab_model(m)
```

### Player Boxscore

```{r}

player <- read.csv("/Users/chejeong/Desktop/player_boxscore.csv")

```

We clean the data for the player boxscore data below.

```{r}



```









